Analysis of the 2014 Passenger Flight Network

### 2020-02-12 paper/main.tex
comparing the patronage estimation based on runway size and a default value to patronage estimation based on neighboring airport's patronage by looking at the flowDeviation distribution and the worst constellation in each cases

@@

In order to enable us to do more meaningful analysis, we enrich the dataset with additional information. We chose the airport patronage and found out, that we can get this data from Wikidata. 
For 1956 of the 3139 airports from the dataset, we can find patronage data from Wikidata. 
We need to distinguish between monthly and yearly values and we extrapolate 2014 data from close years in case no data for the year 2014. 
In order to populate the remaining nodes with reasonable patronages as well, we tried different techniques. 
First we used data about runway surfaces that we found on ourflights.org. We summed up the surfaces of runways for every airport and then calculated the average ratio between total runway surface area and patronage of airports where both data points were present, in order to predict patronage for airports without patronage data but with runway surface data based on this average ratio. For the still remaining >1000 airports without patronage we would then apply a default value that equals the avg of the smallest 1\% of airports with patronage data from Wikidata. 
The runway data turned out to be very unclean and unreliably. Water runways and other quirks were distorting the predictions.

Inspired by Prof. Ikonnikova advice during the presentation we switched to estimating the remaining airport patronages by looking at neighboring airports with patronage data.

\begin{lstlisting}[language=Python, caption=Patronage Estimation]
G = loadEntireNetwork()
pr = -1 #previous remaining
r = 0   #remaining
while r != pr:
  pr = r
  r = 0
  ep = {} #estimated patronages
  for apid in G.nodes:
    node = G.nodes[apid]
    if node["patronage"] == None:
      neighbors = G.neighbors(apid) 
      deg = G.degree(apid)
      sum = 0
      count = 0
      for nid in neighbors: 
        n = G.nodes[nid]
        npatr = n["patronage"]
        ndeg = G.degree(nid) 
        if npatr != None: 
          sum += deg * npatr / ndeg
          count += 1
      if count == 0:
        r += 1
      else:
        avg = sum / count
        ep[apid] = avg 
  for apid in ep: 
    node = G.nodes[apid]
    node["patronage"] = ep[apid]
\end{lstlisting}

This second approach yields much nicer estimations, as we will see after estimating the flows between airports.

@@

The flows are estimated by choosing reasonable initial values and then applying corrections iteratively to approach a situation where the sum of the estimated flows for each airport equal its patronage value.
The average initial values for an edge between airport $p$ and $q$ is chosen as: 
\begin{equation}
avgFlow(pq) = \dfrac{patronage(p) * patronage(q)}{degree(p) * degree(q)}
\end{equation}
Remember that we are handling a multi-edge directed graph here. 
\begin{equation}
E = \{(pql) | \exists\ \text{route from}\ p\ \text{to}\ q\ \text{by}\ l\}
\end{equation}
where $p$ and $q$ are airports and $l$ is an airline.
Let us introduce a function $E(pq)$ that returns all edges between $p$ and $q$ disregarding the direction and operating airline. 
\begin{equation}
E(pq) = \{pql | \exists_{l \in L}:pql \in E \lor qpl \in E\}
\end{equation}
where $L$ is the set of airlines.

And let us introduce the function $E(p)$ that returns all edges from and to an airport $p$ by every airline: 
\begin{equation}
E(p) = \{rsl | (r=p \lor s=p) \land \exists_{l \in L}:rsl \in E \}
\end{equation}
The Edges of the graph are thus triplets of a source airport, destination airport and an airline.
There may be multiple edges between two airports pointing in different directions and/or operated by different airlines.
We could assume an equal distribution of the flows among the edges. But we chose trying to be closer to reality by weighting the flow of an edge between two airports $p$ and $q$, that is operated by airline $l$, by the geometric mean of the number of edges airline $l$ operates at airport $p$ and number of edges airline $l$ operates at airport $q$: 
\begin{equation}
\begin{aligned}
&flow(pql) =\\
={}&avgFlow(pq)  
* noe(pq) 
* weight(pql) 
\end{aligned}
\end{equation}
where $noe(pq)$ is the number of edges between airport $p$ and $q$ and $weight(pql)$ is defined as follows: 
\begin{equation}
\begin{aligned}
&weight(pql) = \dfrac{\sqrt{degree_{l}(p) * degree_{l}(q)}}{pnws(pql)}
\end{aligned}
\end{equation}
where the $pnws(pq)$ (Pre Normalization Weight Sum) is just the sum of the non-normalized weights of all the routes from different airlines to normalize the weights. 
\begin{equation}
\begin{aligned}
&pnws(pq) = \\
={}&\displaystyle\sum_{rsm \in E(pq)}(\sqrt{degree_{m}(r) * degree_{m}(s)})
\end{aligned}
\end{equation}
where $degree_l(p)$ is the degree of airport $p$ in airline $l$'s subgraph, i.e. number of edges from and to airport $p$ operated by airline $l$.

The ratios we choose here will stay intact over the course of the iterations in the next part of the algorithm. 
In every iteration, we traverse all airports and for every airport we calculate the sum of current flows values of all routes. 
\begin{equation}
\begin{aligned}
flowSum(p) = \displaystyle\sum_{pql \in E \cup pu \in E}flow_{pql}
\end{aligned}
\end{equation}
Difference between the patronage and the current sum of passenger flows of incoming and outgoing routes. 
Then, for each airport we calculate the difference between its patronage and the current sum of flows.
\begin{equation}
\begin{aligned}
flowDiff(p) = patronage(p) - flowSum(p) 
\end{aligned}
\end{equation}
With this information given, we can now calculate for every edge $pql$ the change in its flow. 
\begin{equation}
\begin{aligned}
\forall_{pql \in E(p)}: & \\
& \Delta flow(pql) = & \\ 
& ={} \dfrac{flow(pql)}{flowSum(p)}flowDiff(p) \\
& +{} \dfrac{flow(pql)}{flowSum(q)}flowDiff(q) \\
\end{aligned}
\end{equation}
and the more perfect flows for the edges and starting point for the next itaration are
\begin{equation}
\begin{aligned}
\forall_{pql \in E(p)}: & \\
& flow'(pql) = flow_{abo} + \Delta flow(pql)
\end{aligned}
\end{equation}

In order to look at the results of this algorithm we introduce a measure 
\begin{equation}
flowDeviation(p) = ln(flowSum(p)/patronage(p))
\end{equation}
it expresses how well the sum of estimated flows match the patronage, that we know for the airport. If the flows are half the patronage the absolute value of $flowDeviation$ will be the same as when the current flows are twice the actual patronage number. 
We compare the results between estimating the patronages based on the runway sizes Figure~\ref{fig:flow_deviation_runway} and estimating the patronages based on neighboring airports Figure~\ref{fig:flow_deviation}. 
The blue line shows the $flowDeviation$ distribution of the initial flow values.
The other almost totally overlapping lines show the $flowDeviation$ distribution after a couple of iterations.

The bad results of the estimations when estimating patronage values from runways stem from constellations like in Figure~\ref{fig:impossible_flows}, when choosing flows to match the patronage values are impossible.

\begin{figure}
	\centering
	\includegraphics[width=76mm]{imgs/flow_estimation_runway_approach_deviation.png}
	\caption{
Estimated flows when estimating patronage based on the runway size }
	\label{fig:flow_deviation_runway}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=76mm]{imgs/results/flow_estimation_120_iterations.png}
	\caption{
Estimated flows when estimating patronage based on neighboring airport's patronage}
	\label{fig:flow_deviation}
\end{figure}

@@

\section{Basic Network Analysis}

>>>>

% for the equation stuff 
\DeclareMathOperator{\avgFlow}{avgFlow}
\DeclareMathOperator{\patronage}{patronage}
\DeclareMathOperator{\flow}{flow}
\DeclareMathOperator{\weight}{weight}
\DeclareMathOperator{\PNWS}{PNWS}
\DeclareMathOperator{\NOE}{NOE}
\DeclareMathOperator{\degree}{degree}
\DeclareMathOperator{\flowSum}{flowSum}
\DeclareMathOperator{\flowDeviation}{flowDeviation}
\DeclareMathOperator{\flowDiff}{flowDiff}

@@

In order to enable us to do more meaningful analysis, we enrich the dataset with additional information. We chose the airport patronage and found out, that we can get this data from Wikidata. 
For 1956 of the 3139 airports from the dataset, we can find patronage data from Wikidata. Get an impression about the regional differences of data availability by looking at the percentage of airports for which Wikidata yields a patronage in Table~\ref{tab:patr_av_most_airports} and Table~\ref{tab:patr_av_worst}.
\begin{table}[h]
	\centering
	\caption{patronage data availability on Wikidata for countries with the most airports}
	\label{tab:patr_av_most_airports}
	\begin{tabular}{|l|c|c|}
		\hline
		Country & Airports & patronage present \\ \hline
		United States & 539 & 34.1\%\\
		China & 173 & 95.4\%\\
		Canada & 142 & 48.6\%\\
		Brazil & 122 & 93.4\%\\
		Australia & 113 & 77.0\%\\
		Russia & 103 & 96.1\%\\
		India & 70 & 92.9\%\\
		Indonesia & 63 & 60.3\%\\
		Japan & 62 & 82.3\%\\
		Mexico & 56 & 92.9\%\\
		France & 55 & 96.4\%\\ \hline
	\end{tabular}
\end{table}
\begin{table}[h]
	\centering
	\caption{patronage data availability - worst countries}
	\label{tab:patr_av_worst}
	\begin{tabular}{|l|c|c|}
		\hline
		Country & Airports & patronage present \\ \hline
		Venezuela & 23 & 4.3\%\\
		Bahamas & 17 & 5.9\%\\
		Chile & 16 & 6.3\%\\
		Ecuador & 14 & 7.1\%\\
		Ethiopia & 13 & 7.7\%\\ 
		Mozambique & 10 & 10.0\%\\
		Cuba & 9 & 11.1\%\\
		Kenya & 16 & 12.5\%\\
		Namibia & 8 & 12.5\%\\
		Madagascar & 13 & 15.4\%\\
		Algeria & 26 & 19.2\%\\ \hline
	\end{tabular}
\end{table}
We need to distinguish between monthly and yearly values and we extrapolate 2014 data from close years in case no data for the year 2014. 
In order to populate the remaining nodes with reasonable patronages as well, we tried different techniques. 
First we used data about runway surfaces that we found on ourflights.org.

We summed up the surfaces of runways for every airport and then calculated the average ratio between total runway surface area and patronage of airports where both data points were present, in order to predict patronage for airports without patronage data but with runway surface data based on this average ratio. For the still remaining >1000 airports without patronage we would then apply a default value that equals the average of the smallest 1\% of airports with patronage data from Wikidata. 
The runway data turned out to be very unclean and unreliably. Water runways and other quirks were distorting the predictions.

Inspired by Prof. Ikonnikova advice during the presentation we switched to estimating the remaining airport patronages by looking at neighboring airports with patronage data. The Algorithm can be seen in Listing~\ref{lst:patronage_estimation_algorithm}.
\begin{lstlisting}[language=Python, caption={Patronage estimation algorithm, that using neighboring airport's patronage data}, label=lst:patronage_estimation_algorithm]
G = loadEntireNetwork()
pr = -1 #previous remaining
r = 0   #remaining
while r != pr:
  pr = r
  r = 0
  ep = {} #estimated patronages
  for apid in G.nodes:
    node = G.nodes[apid]
    if node["patronage"] == None:
      neighbors = G.neighbors(apid) 
      deg = G.degree(apid)
      sum = 0
      count = 0
      for nid in neighbors: 
        n = G.nodes[nid]
        npatr = n["patronage"]
        ndeg = G.degree(nid) 
        if npatr != None: 
          sum += deg * npatr / ndeg
          count += 1
      if count == 0:
        r += 1
      else:
        avg = sum / count
        ep[apid] = avg 
  for apid in ep: 
    node = G.nodes[apid]
    node["patronage"] = ep[apid]
\end{lstlisting}
This second approach yields much nicer estimations, as we will see after estimating the flows between airports.

@@

The flows are estimated by choosing reasonable initial values and then applying corrections iteratively to approach a situation where the sum of the estimated flows for each airport equal its patronage value.
The average initial values for an edge between airport $p$ and $q$ is chosen as:

\begin{equation}
\avgFlow(pq) = \dfrac{\patronage(p) * \patronage(q)}{\degree(p) * \degree(q)}
\end{equation}
Remember that we are handling a multi-edge directed graph here. 
\begin{equation}
E = \{(pql) | \exists\ \text{route from}\ p\ \text{to}\ q\ \text{by}\ l\}
\end{equation}
where $p$ and $q$ are airports and $l$ is an airline.
Let us introduce a function $E(pq)$ that returns all edges between $p$ and $q$ disregarding the direction and operating airline. 
\begin{equation}
E(pq) = \{pql | \exists_{l \in L}:pql \in E \lor qpl \in E\}
\end{equation}
where $L$ is the set of airlines.

And let us introduce the function $E(p)$ that returns all edges from and to an airport $p$ by every airline: 
\begin{equation}
E(p) = \{rsl | (r=p \lor s=p) \land \exists_{l \in L}:rsl \in E \}
\end{equation}
The Edges of the graph are thus triplets of a source airport, destination airport and an airline.
There may be multiple edges between two airports pointing in different directions and/or operated by different airlines.
We could assume an equal distribution of the flows among the edges. But we chose trying to be closer to reality by weighting the flow of an edge between two airports $p$ and $q$, that is operated by airline $l$, by the geometric mean of the number of edges airline $l$ operates at airport $p$ and number of edges airline $l$ operates at airport $q$: 
\begin{equation}
\begin{aligned}
&\flow(pql) =\\
={}&\avgFlow(pq)  
* \NOE(pq) 
* \weight(pql) 
\end{aligned}
\end{equation}
where $\NOE(pq)$ is the number of edges between airport $p$ and $q$ and $\weight(pql)$ is defined as follows: 
\begin{equation}
\begin{aligned}
&\weight(pql) = \dfrac{\sqrt{\degree_{l}(p) * \degree_{l}(q)}}{\PNWS(pql)}
\end{aligned}
\end{equation}
where the $\PNWS(pq)$ (Pre Normalization Weight Sum) is just the sum of the non-normalized weights of all the routes from different airlines used to normalize the weights. 
\begin{equation}
\begin{aligned}
&\PNWS(pq) = \\
={}&\displaystyle\sum_{rsm \in E(pq)}(\sqrt{\degree_{m}(r) * \degree_{m}(s)})
\end{aligned}
\end{equation}
where $\degree_l(p)$ is the degree of airport $p$ in airline $l$'s subgraph, i.e. number of edges from and to airport $p$ operated by airline $l$.

The ratios we choose here will stay intact over the course of the iterations in the next part of the algorithm.
In every iteration, we traverse all airports and for every airport we calculate the sum of current flows values of all routes. 
\begin{equation}
\begin{aligned}
\flowSum(p) = \displaystyle\sum_{pql \in E \cup pu \in E}\flow_{pql}
\end{aligned}
\end{equation}
Difference between the patronage and the current sum of passenger flows of incoming and outgoing routes. 
Then, for each airport we calculate the difference between its patronage and the current sum of flows.
\begin{equation}
\begin{aligned}
\flowDiff(p) = \patronage(p) - \flowSum(p) 
\end{aligned}
\end{equation}
With this information given, we can now calculate for every edge $pql$ the change in its flow. 
\begin{equation}
\begin{aligned}
\forall_{pql \in E(p)}: & \\
& \Delta \flow(pql) = & \\ 
& ={} \dfrac{\flow(pql)}{\flowSum(p)}\flowDiff(p) \\
& +{} \dfrac{\flow(pql)}{\flowSum(q)}\flowDiff(q) \\
\end{aligned}
\end{equation}
and the more perfect flows for the edges and starting point for the next itaration are
\begin{equation}
\begin{aligned}
\forall_{pql \in E(p)}: & \\
& \flow'(pql) = \flow_{abo} + \Delta \flow(pql)
\end{aligned}
\end{equation}

In order to look at the results of this algorithm we introduce a measure 
\begin{equation}
\flowDeviation(p) = ln\left(\dfrac{\flowSum(p)}{\patronage(p)}\right)
\end{equation}
it expresses how well the sum of estimated flows match the patronage, that we know for the airport. If the flows are half the patronage the absolute value of $\flowDeviation$ will be the same as when the current flows are twice the actual patronage number. 
We compare the results between estimating the patronages based on the runway sizes Figure~\ref{fig:flow_deviation_runway} and estimating the patronages based on neighboring airports Figure~\ref{fig:flow_deviation}. 
The blue line shows the $\flowDeviation$ distribution of the initial flow values.
The other lines show the $\flowDeviation$ distribution after a the first couple of iterations and after the final iteration. 
In the case of patronage estimations based on neighboring airports, we see that the $\flowDeviation$ distribution after the 120th iteration comes very close to $y=0$ which is also reflected by the $sum(abs)$ in the legend which is the sum of the absolute value of $\flowDeviation$ over all airports which is $0.94$ in this case.

The bad results of the estimations when estimating patronage values from runways stem from constellations like in Figure~\ref{fig:impossible_flows}, when choosing flows to match the patronage values are impossible. 
The constellation that accounts for $>0.88$ of the $\flowDeviation$ sum $0.94$ in the scenario where patronages are estimated based on neighboring airports is shown in Figure~\ref{fig:worst_flow_fit}. 
The patronage of both airports comes from Wikidata. After further investigation it turns out to be probable, that there should be a connection between Arvidsjaur Airport and Stockholm-Arlanda Airport, which is not present in the data set.

\begin{figure}[h]
	\centering
	\includegraphics[width=76mm]{imgs/flow_estimation_runway_approach_deviation.png}
	\caption{
Estimated flows when estimating patronage based on the runway size and a default value}
	\label{fig:flow_deviation_runway}
\end{figure}

\begin{figure}[h]
	\centering
	\includegraphics[width=76mm]{imgs/results/flow_estimation_120_iterations.png}
	\caption{
Estimated flows when estimating patronage based on neighboring airport's patronage}
	\label{fig:flow_deviation}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\includegraphics[width=76mm]{imgs/runway-based-estimation-problem.png}
	\caption{Worst situation in the scenario, where patronages are estimated based on runways and a default value: 
For Qaanaaq Airport there was no patronage data from Wikidata and no runway information. That's why it ended up with the default patronage value 1523. In this constallation it is impossible to choose passenger flows that even remotely satisfy the the patronages of airports with ids 10 and 5446}
	\label{fig:impossible_flows}
\end{figure}

\begin{figure}[h]

	\centering
	\includegraphics[width=76mm]{imgs/results/ego_graph_730.png}
	\caption{Worst situation in the scenario, where patronage is based on neighboring airport's patronage: 
Arvidsjaur Airport is only connected to Lycksele Airport but has a more than twice as high patronage than Lycksele Airport}
	\label{fig:worst_flow_fit}
\end{figure}
\section{Basic Network Analysis}
